### 2017-03-14 ThesisText/data.tex
Wrote data chapter

\chapter{Data}

@@

In order to explore the data using the bash tools (cut, grep, etc) and tools
like Python with Pandas, they were exported to a CSV format. Some
transformations may be applied to these files in other to clean useless data
points or make the data easier to process.

To begin, each sheet is analysed individually. Conclusions on each list are
registered in the fist section of this chapter. After analysing the original
data, the second section of the chapter asserts which entities and objects are
present in this data. Finally, the third section lists the datasets which
represent those entities and objects.

\subsection{Sheet 1 - \texttt{UCs\_SIIUE\_Moodle.csv}}

This sheet lists a few courses (referred to in their native name as
\textit{Unidades Curriculares}). Some courses have a Moodle page while others
do not.

The sheets contain four non blank columns. The first and second column are a
list of courses represented has their Course Id and Moodle Id, which is
different.

The third column is just a calculated copy of the first. It doesn't add
anything and it might have been a loose processing step at some point.

The fourth column contains a list of course with no Moodle and therefore no
Moodle Id.

Only the following columns are relevant:

\begin{itemize}
    \item \texttt{Cod UC} - Course's code
    \item \texttt{No moodle} - Course's code in the Moodle
\end{itemize}

\subsection{Sheet 2 - \texttt{complete\_rate.csv}}

This sheet contain a listing of the completion rate of each course. The list
has tree columns. A column for course id in the Moodle (\texttt{c\_moodleid}),
the number of students in that course (\texttt{n\_students}), and the
completion rate of those courses (\texttt{complete\_rate}).

So a course with 100 students having a completion rate of 75\% means that 75
students actually completed that course.

\subsection{Sheet 3 - \texttt{completed.csv}}

Contains a listing of which student was approved at which course. The sheet has
three columns. The first column is the course id of the Moodle
(\texttt{c\_moodleid}), the second is the user id of the Moodle
(\texttt{u\_moodleid}), and the third is a column (\texttt{completed}) which
has the value 1 if the student completed that course, or 0 if not.

The user id may be assigned to a professor, but in this sheet every user id is
of a student.

\subsection{Sheet 4 - \texttt{Notas UCS SIIUE.csv}}

There is a lot of information in this sheet. Basically, this sheet contains a
big list. That list has a course, a student, and the student's results on that
course. The results contain the grade, if available, if the student was
approved or not, and if not, why? Did he failed the evaluations or simply
missed them?

Some statistics are calculated in a table to the right of this sheet. Those
calculated values are useful, but should be recalculated elsewhere in a later
time if necessary.

\subsection{Sheet 5 - \texttt{Profile}}

This sheet contains some analysis made to the data. To begin with, there are
the following columns:

\begin{itemize}
    \item \texttt{name\_c} - Name of the course
    \item \texttt{name\_s} - Name of the student
    \item \texttt{id\_c} - Id of the course
    \item \texttt{n\_resrc\_c} - Number of resources (files, urls, etc)
    \item \texttt{n\_actv\_c} - Number of activities (forums, quizzes, etc)
    \item \texttt{id\_s} - Student's id
    \item \texttt{n\_cons\_s} - Number of views
    \item \texttt{n\_subm\_s} - Number of submissions
    \item \texttt{result\_s} - Results of that student in that course
\end{itemize}

In order for a student to get approved, some criteria must be followed. That
criteria is described in the sheet itself.

\subsection{Sheet 6 - \texttt{ResultadosUCMoodle}}

This sheet contains a lot of information about the courses. To begin with, we
get a relation between the three ways of identifying a course in the data. A
course is identified by the Course's Code (\texttt{Cod Siiue}), Course's Code
in the Moodle (\texttt{Cod Moodle}), or by a course id (\texttt{course}). The
course id is an integer number, while the codes contain information about the
department.

The department (\texttt{Dep}) is calculate from the course's code (\texttt{Cod
Siiue}) by taking the first three characters from that course. So a course with
code \texttt{INF7185\_1} will be from the department \texttt{INF}.

The sheet presents the first and last register of an activity in the Moodle,
and also the interval in days from the first activity to the last. The fields
are \texttt{min}, \texttt{max}, and \texttt{int}.

The sheet finally specifies the number of students who enrolled in a course and
the number of students who were approved on that course.

\begin{itemize}
    \item \texttt{course} - Course's id
    \item \texttt{Cod Moodle} - Course's Code in the Moodle
    \item \texttt{Cod Siiue} - Course's Code
    \item \texttt{min} - First activity of course
    \item \texttt{max} - Last activity of course
    \item \texttt{int} - Interval between first and last activity of course
    \item \texttt{actsub} - Number of Submission Activities(Same value ad above)
    \item \texttt{actcons} - Number of views (Same as resources)
    \item \texttt{nactsubd} - Number of different activities
    \item \texttt{consal} - Student views
    \item \texttt{consdoc} - Professor views
    \item \texttt{subal} - Student submission
    \item \texttt{subdoc} - Professor submission
    \item \texttt{nalu} - Number of students in the course
    \item \texttt{ndoc} - Number of professors in the course
    \item \texttt{InscSIIUE} - Number of students who enrolled
    \item \texttt{Aprov} - Number of students who got approved
    \item \texttt{Dep} - Department of the course
\end{itemize}

\subsection{Sheet 7 - \texttt{Correlação}}

This is a sheet that implements the calculation of correlation between the
variables of the previous sheet \texttt{ResultadosUCMoodle}. The data it self
is coppied from the previous sheet and a new table to the right of the first
table is added with the correlation itself.

The data in here doesn't need to be preprocessed to a new sheet, but the notion
of correlation, which can be calculated using alternative methods, may be used
later.

\subsection{Sheet 8 - \texttt{Alunos SIIUE E-L}}

There are a few issues with this sheet. First of all, the sheet appears to have
two tables which are not really related, maybe? Anyway. The first table has a
student number, his name, and what appears to be a count of it's enrollments.
It also has a calculated fields stating his/her's first and last name and
another column with the students number which is a calculated copy of the
first.

The second table has Moodle usernames. A first and last name, another count
(which I assume is a count of enrollments but I can't be sure because the field
simply states ``count'').

Oh! The second number in the first table is used for a calculated value in the
second table.

So the last three columns of the second table are all calculated values. The
first appears to fetch a number from the first. It goes like this. If the name
of the student in the second table is equal to the first and last name of the
first, then fetch that number. Else, it is an error.

Next, we get a value, is the username in the Moodle contains a number. If not,
we get an error.

Lastly. We try to get the first calculated number (the fourth column of the
second table) if it exists. If it doesn't, we try to get the second calculated
value (the fifth column of the second table). If that doesn't exist too, we
write 0.

Up to this point I'm not sure what the purpose of this sheet is. I believe it
gives me a relation between a student in SIIUE and in Moodle, but I have
serious doubts about the correctiveness of that relation.

After examining the sheet a second, I now believe that the first table has a
listing o students and their SIIUE internal id or student number. The second
table simply has usernames of Moodle. So the second table has students and
professors, while the first one has students only.

About the counts. The count of the first table is indeed a number of courses
where the students is enrolled in. The count of the second table is the total
number of log entries which is a copy (a copy-paste values only copy) of the
information in the \texttt{AlunosMoodle} sheet.

I don't think this sheet alone is much useful. But it did give insight on how
this data repository is organized.

\subsection{Sheet 9 - \texttt{AlunosMoodle}}

These are the actual logs of the accesses to Moodle. This is a big table that
contains the following fields:

\begin{itemize}
    \item \texttt{c\_id} - Course ID in the Moodle
    \item \texttt{c\_cod} - Course ID
    \item \texttt{c\_role} - Role of the user
    \item \texttt{c\_user} - User's code
    \item \texttt{c\_nome} - User's name
    \item \texttt{c\_uid} - User's ID in the Moodle
    \item \texttt{crud} - \gls{crud} activity
    \item \texttt{semana} - Number of the week in which the events happened
    \item \texttt{n} - Number of time the event happened on that week
    \item \texttt{numAluno} - Number of the student, which is calculated from
        elsewhere, even when the user is not a student
    \item \texttt{SNot} - If the student was approved, or failed, or something
        else
\end{itemize}

Basically, we get a list of entries specifying activities in Moodle. Each entry
contains information on the course, the user, and the activity itself.

For the course information we have course id and course code (Moodle code).

For user, we get the user's role, username, actual name of user, and some
number.

Finally, we get what kind of CRUD activity it was (if create activity, delete,
etc), the week in which said activity occurred, and number of times the
activity occurred in that week. The activities are aggregated by type and week.

This table then contains the fields \texttt{numAluno} and \texttt{SNot}. These
are calculate fields. The first attempts to take a student's number from the
username, if possible. The second calculates if that particular student was
approved in the course for which the entries refers to. Both of these fields
are useless.

There is a second table. This table gives a username, actual name, and total
number of activities of that user in the log. It is useful, but it should be in
it's own sheet. The fields of that table have no name, but in order to be
consistent with the table on the left, we get:

\begin{itemize}
    \item \texttt{c\_user}
    \item \texttt{c\_nome}
    \item \texttt{count}
\end{itemize}

\subsection{Sheet 10 - \texttt{CalculosAlunos}}

I'm not sure what his sheet is. It has a few values copied from other tables,
has ``\textit{Calculos}'', on the name, but it doesn't really calculate
anything. All the values are literal.

It may be discarded.

\subsection{Sheet 11 - \texttt{export\_weekly}}

This table seems to have a line for each pain of course and user ids. It then
has a few columns of log activities. For each intersection of pair of course
and user ids we get a 1 or 0 for one of the columns. Presumably, we get a true,
that is, 1, if a user A made that activity in course B in that week.

This isn't really useful because the same can be easily calculated using other
tools when needed.

\subsection{Sheet 12 - \texttt{export\_activity}}

This table seems to have a few more calculations, but I don't fully understand
them. It has at least a calculation of a standard deviation, and little else.

Again, this isn't really useful because the same can be easily calculated using
other tools when needed.

% =============================================================================
\section{Objects, Entities, and Features}
\label{sec:objs_ents_feat}

After a first analysis of the data, the objects in the data are identified.
These objects are the courses, students, results, etc. Data mining tasks are
applied over these objects which are described by their features.

@@

In the data, a course is identified by one of three unique ids. They are, the
Course ID, SIIUE's Course's Code, and Moodle Course's Code. The Course id is
simply a unique positive integer number which is probably assigned as an
incremental value when the courses where created in some database.

@@

Table~\ref{tab:course_code_example} shows an example of the three ids. Notice how \texttt{CourseCodeMoodle} is either the same as \texttt{CourseCodeSiiue} or has just minimal differences (in this case there is an underscore follwed by a number).

@@

A course also has a regime, a semester, a season, and a type. These features
are enumerated types because they have a limited domain. For example, the
semester may either be ``\textit{Par}'' or ``\textit{Ímpar}'' (Even or Odd),
the season may be ``\textit{Normal}'' or ``\textit{Special}''.

@@

Some courses have no Moodle. They are therefore useless for our these
experiments.

Some statistics are taken from the courses data as shown in
table~\ref{tab:courses_stats}.

The domain of the enumerated fields are listed in table~\ref{tab:courses_enum}.
The field ``type'' only has the domain ``\textit{Normal}'', so it is a field
which is useless for further analysis. The field ``Semester'' has a missing
values. In the actual data there is only one course which doesn't have a value
for this field.

@@

    \begin{tabular}{| l | r |}
        \hline
        Number of Courses     & 77 \\ \hline
        Number of Departments & 9  \\ \hline
        Number of Degrees     & 11 \\ \hline
    \end{tabular}

    \caption
        [Statistics on the Courses]
        {Statistics on the Courses.}

    \label{tab:courses_stats}
\end{table}

    \begin{tabular}{| l | l |}
        \hline
        Regime   & \{ A, O, S \}                                    \\ \hline
        Semester & \{ ``\textit{Par}'', ``\textit{Ímpar}'', n/a \}  \\ \hline
        Season   & \{ ``\textit{Normal}'', ``\textit{Especial}'' \} \\ \hline
        Type     & \{ ``\textit{Normal}'' \}                        \\ \hline
    \end{tabular}

@@

Courses have different values for credits. The statistic program
\texttt{stat\_001} lists number of courses with a given value for credits.
Table~\ref{tab:courses_credits} shows that count.

    \begin{tabular}{| l | r |}
        \hline
        Credits & Number of Courses \\ \hline
        1       & 3                 \\ \hline
        2       & 3                 \\ \hline
        3       & 3                 \\ \hline
        5       & 6                 \\ \hline
        6       & 46                \\ \hline
        7.5     & 12                \\ \hline
        12      & 2                 \\ \hline
        15      & 2                 \\ \hline
    \end{tabular}

    \caption
        [Number of courses group by credits]
        {Number of courses group by credits.}

    \label{tab:courses_credits}
\end{table}

@@

The values for the \textit{StudentType} field are:

\begin{itemize}
    \item \textit{Normal}
    \item \textit{Interno}
\end{itemize}

Professors have only the features listed in table~\ref{tab:professor_features}.
The data contains a field called \textit{Role} which states if that Moodle user
is a professor or a student. Supposedly, if the user is a professor the role
has a value of 5. If it is a student, it has a value of 3. However, we see
entries for professors with a role value of 3 and vice versa. It is unknown if
this role actually state the if a user is a professor or student or something
else.

@@

Table~\ref{tab:moodle_users_stats} shows some stats regarding Moodle users.

    \begin{tabular}{| l | r |}
        \hline
        Number of Students   & 145 \\ \hline
        Number of Professors & 87  \\ \hline
        Total                & 232 \\ \hline
    \end{tabular}

    \label{tab:moodle_users_stats}
\end{table}

@@

Table~\ref{tab:results_stats} shows some statistics on the results data.

    \begin{tabular}{| l | r |}
        \hline
        Number of results           & 716 \\ \hline
        Number of non-empty results & 57  \\ \hline
        Number of approved students & 334 \\ \hline
    \end{tabular}

    \caption
        [Statistics on the Results]
        {Statistics on the Results.}

    \label{tab:results_stats}
\end{table}

@@

% =============================================================================
\section{Preprocessed Datasets}

@@

The data stored in CSV files from which any statistics are calculated and
experiments are executed. The data is also organized following a relational
paradigm, even if the data is not stored in a relational database.

The following sections describe which datasets store the data of the various
objects as described in section~\ref{sec:objs_ents_feat}. In the end of this
section, figure\ref{fig:er} shows the relational model of the various datasets.

Data on the courses is mostly store in the following dataset files:

The \texttt{CoursesStudents} dataset contains one entry for each course. Each
entry states the total number of students of that course, the total number of
approved students, and the approval ration for that course. This information
would easily be calculated from other datasets in this repository, but it is
kept here since it was present in the original datasets.

The following three datasets store information on the Moodle users:

\begin{itemize}
    \item \texttt{MoodleUsers}
    \item \texttt{Students}
\end{itemize}

Each Moodle user has a username, an actual name, and a role. That information
is stored in \texttt{MoodleUsers}. Students have the fields stated in
section~\ref{ref:moodle_users}, and those are stored in the \texttt{Students}
dataset.

\subsection{Results and Profiling Datasets}

Results are mostly stored in the \texttt{Results} dataset, but some other
information is stored in \texttt{Profile}.

The results consist of the fields stated in section~\ref{sec:results}.

\TODO{Talk about the profile dataset}

@@

>>>>

\chapter{University of Évora's Moodle Data}

@@

In order to explore the data using the bash tools (cut, grep, etc) and tools
like Python with Pandas, they were exported to a CSV format. Some
transformations were applied to these files in other to clean useless data
points or make the data easier to process.

The provided sheets contain information of courses, students, their activities
in Moodle, and their results. The data comes from Moodle databases and from the
SIIUE system. Both systems are integrated, making Moodle plus SIIUE the
University's own~\gls{lms}.

The sheets are the following:

\begin{itemize}
    \item \texttt{UCs\_SIIUE\_Moodle.csv}
    \item \texttt{complete\_rate.csv}
    \item \texttt{completed.csv}
    \item \texttt{Notas UCS SIIUE.csv}
    \item \texttt{Profile}
    \item \texttt{ResultadosUCMoodle}
    \item \texttt{Correlação}
    \item \texttt{Alunos SIIUE E-L}
    \item \texttt{AlunosMoodle}
    \item \texttt{export\_weekly}
    \item \texttt{export\_activity}
\end{itemize}

Some sheets contained information that was the result of calculations from the
established data. Analysis like these were removed from the data for two
reasons. First, some analysis are not useful so them being in the data only
makes the data more complex. By removing such analysis we get a simpler data
model. Second, even if some analysis are in fact useful, they are still easily
reproduced by the Python tools, with the difference that they are only
reproduced when needed for specific experiments.

Some analysis were made and are described in section~\ref{set:stats}

The sheets that remained after removing analysis data are:

\begin{itemize}
    \item \texttt{UCs\_SIIUE\_Moodle.csv}
    \item \texttt{Notas UCS SIIUE.csv}
    \item \texttt{Profile}
    \item \texttt{ResultadosUCMoodle}
    \item \texttt{Alunos SIIUE E-L}
    \item \texttt{AlunosMoodle}
\end{itemize}

The rest of the sheets contain the objects which are going to be used in the
Data Mining tasks. Those objects and their features are fairly unorganized in
the original data, so an analysis on of the available features had to be done.
Section~\ref{sec:objs_feat} states the objects found in the data and details
their features. Section~\ref{sec:preproc_dataset} shows how the final dataset
is organized.

% =============================================================================
\section{Objects and Features}
\label{sec:objs_feat}

After a first analysis of the data, the objects are identified. These objects
are courses, students, results, etc. Data mining tasks are applied over these
objects which are described by their features.

@@

In the data, a course is identified by one of three unique ids. They are, the
Course Id, SIIUE's Course's Code, and Moodle Course's Code. The Course Id is
simply a unique positive integer number which is probably assigned as an
incremental value when the courses where created in some database.

@@

Table~\ref{tab:course_code_example} shows an example of the three ids. Notice
how \texttt{CourseCodeMoodle} is either the same as \texttt{CourseCodeSiiue} or
has just minimal differences (in this case there is an underscore followed by a
number).

@@

A course also has a regime, a semester, a season, and a type. These features
are enumerated types because they have a limited domain. For example, the
semester may either be ``\textit{Par}'' or ``\textit{Ímpar}'' (Even or Odd).
The type only has one possible value which is \texttt{Normal}. Despite this,
the field was kept in the data for completeness. Table~\ref{tab:courses_enum}
shows the domain of each enumerated type.

@@

Some courses have no Moodle and, as such, have no other associated information.
They are therefore useless for these experiments and were simply removed.

@@

    \begin{tabular}{| l | l |}
        \hline
        Regime   & \{ A, O, S \}                            \\ \hline
        Semester & \{ \textit{Par}, \textit{Ímpar}, n/a \}  \\ \hline
        Season   & \{ \textit{Normal}, \textit{Especial} \} \\ \hline
        Type     & \{ \textit{Normal} \}                    \\ \hline
    \end{tabular}

@@

@@

The Name field contains the actual name of the student. This information is
useless for any analysis, but the name was kept for completeness.

The values for the \textit{StudentType} field are \textit{Normal} or
\textit{Interno} (Normal or Internal).

Professors have only the features listed in table~\ref{tab:professor_features}.
The data contains a field called \textit{Role} which states if that Moodle user
is a professor or a student. If the user is a professor the role has a value of
5. If it is a student, it has a value of 3. A few users with a value of 4 are
present. These users have the role of editor, but the Moodle logs only have 11
entries for a user of this role, so they are simply discarded.

@@

@@

\subsection{Profile}
\label{sec:profile}

Profile objects are objects that, for each pair of course and student, state
the following indicators:

\begin{itemize}
    \item \texttt{NumberOfResources}
    \item \texttt{NumberOfActivities}
    \item \texttt{NumberOfViews}
    \item \texttt{NumberOfSubmissions}
\end{itemize}

A resource in this context is any file or link that a professor may have placed
in a Moodle page. Profile objects state the number of resources which were seen
by a particular student in a particular course.

An activity is any interaction in a Moodle forum or quizzed. A view is a view
of any page, resource, etc. A submission is when a students submits a project.

@@

% =============================================================================
\section{Preprocessed Datasets}
\label{sec:preproc_dataset}

@@

The data stored in CSV files from which any statistics are calculated and
experiments are executed. The data is also organized following a relational
paradigm.~\footnote{Despite the fact that the data is not kept in a Relational
Database, having it organized in a relational paradigm simplifies the process
of working with it using the Python tools}

The following sections describe which datasets store the data of the various
objects as described in section~\ref{sec:objs_feat}. In the end of this
section, figure~\ref{fig:er} shows the relational model of the various datasets.

@@

Data on the courses is store in the following dataset files:

The \texttt{CoursesStudents} dataset contains one entry for each course. Each
entry states the total number of students of that course, the total number of
approved students, and the approval ration for that course. This information
would easily be calculated from other datasets in this repository, but it is
kept here since it was present in the original datasets as a non calculated
field.

Each Moodle user has a username, an actual name, and a role. That information
is stored in \texttt{MoodleUsers}. This dataset contains users which are both
professors and students. They are identified by their role.

Students have the fields stated in section~\ref{ref:moodle_users}, and those
are stored in the \texttt{Students} dataset.

\subsection{Results}

Results are stored in the \texttt{Results} dataset. The results consist of the
fields stated in table~\ref{tab:results_fetures} in section~\ref{sec:results}.

\subsection{Profiling Datasets}

The \texttt{Profile} dataset contains the information stated in
section~\ref{sec:profile}. The fields are shown in
table~\ref{tab:profile_fields}.

    \begin{tabular}{l l}
        Feature                      & Data Type \\ \hline
        \texttt{CourseCodeMoodle}    & String    \\
        \texttt{CourseId}            & String    \\
        \texttt{StudentsId}          & Number    \\
        \texttt{NumberOfResources}   & Number    \\
        \texttt{NumberOfActivities}  & Number    \\
        \texttt{NumberOfViews}       & Number    \\
        \texttt{NumberOfSubmissions} & Number    \\
    \end{tabular}

    \caption
        [Profile features]
        {Profile features.}

    \label{tab:profile_features}
\end{table}
